\textbf{Machine learning classification by fitting amplicon sequences to existing OTUs}

### 2023-06-15 submission/second_revision/manuscript.tex
second revision documents

Machine learning classification using the gut microbiome relies on
assigning 16S rRNA gene sequences into operational taxonomic units
(OTUs) to quantify microbial composition. OTU abundances are then used
to train a classification model that can be applied to classify new
samples. The standard approaches to clustering sequences include
reference-based and \emph{de novo} clustering. Reference-based
clustering requires a well-curated reference database that may not exist
for all systems. \emph{De novo} clustering tends to produce higher
quality OTU assignments than reference-based, but clusters depend on the
sequences in the dataset and therefore OTU assignments will change when
new samples are sequenced. This lack of stability complicates machine
learning classification since new sequences must be reclustered with the
old data and the model retrained with the new OTU assignments. The
OptiFit algorithm addresses these issues by fitting new sequences into
existing OTUs. While OptiFit produces high quality OTU clusters, it is
unclear whether this method for fitting new sequence data into existing
OTUs will impact the performance of classification models trained with
the older data. We used OptiFit to cluster sequences into existing OTUs
and evaluated model performance in classifying a dataset containing
samples from patients with and without colonic screen relevant neoplasia
(SRN). We compared the performance of this model to standard methods
including \emph{de novo} and database-reference-based clustering. We
found that using OptiFit performed as well or better in classifying
SRNs. OptiFit can streamline the process of classifying new samples by
avoiding the need to retrain models using reclustered sequences.

@@

There is great potential for using microbiome data to aid in diagnosis.
A challenge with OTU-based classification models is that 16S rRNA gene
sequences are often assigned to OTUs based on similarity to other
sequences in the dataset. If data are generated from new patients, the
old and new sequences must all be reassigned to OTUs and the
classification model retrained. Yet there is a desire to have a single,
validated model that can be widely deployed. To overcome this obstacle,
we applied the OptiFit clustering algorithm to fit new sequence data to
existing OTUs allowing for reuse of the model. A random forest model
implemented using OptiFit performed as well as the traditional reassign
and retrain approach. This result shows that it is possible to train and
apply machine learning models based on OTU relative abundance data that
do not require retraining or the use of a reference database.

@@

There is increasing evidence for an association between the composition
of the gut microbiome and a variety of diseases, such as crohn's disease
and colorectal cancer (\protect\hyperlink{ref-Morgan2012}{1},
\protect\hyperlink{ref-Sobhani2011}{2}). There is great potential to
diagnose disease with gut microbiome sequence data and machine learning.
Taxonomic composition of microbial communities can be assessed using
amplicon sequencing of the 16S rRNA gene, which is the input to
classification models. Analysis of 16S rRNA gene sequence data generally
relies on assigning sequences into operational taxonomic units (OTUs).
The process of OTU clustering can either be reference-based or \emph{de
novo}. The quality of OTUs generated with reference-based clustering is
generally poor compared to those generated with \emph{de novo}
clustering (\protect\hyperlink{ref-westcott2015}{3}). While \emph{de
novo} clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the sequences within the dataset and the addition of
new data has the potential to redefine OTU cluster composition. The
unstable nature of \emph{de novo} OTU clustering complicates deployment
of machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool \emph{et al.}
introduced OptiFit, a method for fitting new sequence data into existing
OTUs (\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit can
effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on classification
performance. Here, we tested the ability of OptiFit to cluster new
sequence data into existing OTU clusters for the purpose of classifying
disease based on gut microbiome composition.

We compared the ability of several approaches for assigning 16S rRNA
gene sequences to OTUs including, \emph{de novo} and reference-based
clustering. For reference-based clustering, we used closed-reference
clustering to a public database (database-reference-based) and to OTUs
generated from a subset of the samples (self-reference-based). To test
how the model performance compared between these approaches, we used a
publicly available dataset of 16S rRNA gene sequences from stool samples
of healthy subjects (n = 226) as well as subjects with screen-relevant
neoplasia (SRN) consisting of advanced adenoma and carcinoma (n = 229)
(\protect\hyperlink{ref-baxter2016}{5}). For the \emph{de novo}
workflows, all the 16S rRNA sequence data was clustered into OTUs. The
OTU clustering was conducted using two common algorithms: 1) the
OptiClust algorithm in mothur (\protect\hyperlink{ref-westcott2017}{6})
and 2) the VSEARCH algorithm used in QIIME2
(\protect\hyperlink{ref-rognes2016}{7},
\protect\hyperlink{ref-bolyen2019}{8}). For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). We also conducted reference-based OTU
clustering using OptiFit to fit the sequence data into OTUs based on the
greengenes reference database. To compare with another commonly used
method, we also used the VSEARCH algorithm to fit the sequence data to
the greengenes reference (Figure 1B). In the OptiFit self-reference
workflow, the data was split into a training and a testing set. The
training set was clustered into OTUs and used to train a classification
model. The OptiFit algorithm was used to fit sequence data of samples
not part of the original dataset into the existing OTUs, and used the
same model to classify the samples (Figure 1C). For each of the
workflows the process was repeated for 100 random splits of the data to
account for variation caused by the choice of the random number
generator seed.

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is a metric
used to measure OTU cluster quality based on the similarity of all pairs
of sequences and whether they are appropriately clustered or not
(\protect\hyperlink{ref-westcott2015}{3}). MCC scores range between
negative one and one, and measure how well clustering assignment
correlates with the distance between sequences. To ensure that OptiFit
appropriately integrated new sequence data into the existing OTUs, we
expected the MCC scores produced by the OptiFit workflow to be similar
to that of \emph{de novo} clustering using the OptiClust algorithm. In
the OptiFit workflow the test data was fit to the clustered training
data for each of the 100 data splits resulting in an MCC score for each
split of the data. In the remaining workflows, the data was only
clustered once and then split into the training and testing sets
resulting in a single MCC score for each method. Indeed, the MCC scores
were similar between the OptiClust \emph{de novo} (MCC = 0.884) and
OptiFit self-reference workflows (average MCC = 0.879, standard
deviation = 0.002). Consistent with prior findings, the reference-based
methods produced lower MCC scores (OptiFit Greengenes MCC = 0.786;
VSEARCH Greengenes MCC = 0.531) than the \emph{de novo} methods
(OptiClust \emph{de novo} MCC = 0.884; VSEARCH \emph{de novo} MCC =
0.641) (\protect\hyperlink{ref-sovacool2022}{4}). Another metric we
examined for the OptiFit workflow was the fraction of sequences from the
test set that mapped to the reference OTUs. Since sequences that did not
map to reference OTUs were eliminated, if a high percentage of reads did
not map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average 99.8\% (standard deviation = 0.68\%) of sequences in
the subsampled test set mapped to the reference OTUs. This number is
higher than the average fraction of reads mapped in the OptiFit
Greengenes workflow ( 96.8\% +/- 3.5). These results indicate that the
OptiFit self-reference method performed as well as the OptiClust
\emph{de novo} method and better than using an external database.

We next assessed model performance using OTU relative abundances from
the training data from the workflows to train a model to predict SRNs
and used the model on the held out data. Using the predicted and actual
diagnosis classification, we calculated the area under the receiver
operating characteristic curve (AUROC) for each data split. During
cross-validation (CV) training, the performance of the OptiFit
self-reference and OptiClust \emph{de novo} models were not
significantly different (p-value = 0.066; Figure 2A), while performance
for both VSEARCH methods was significantly lower than the OptiClust
\emph{de novo}, OptiFit self, and OptiFit Greengenes methods (p-values
\textless{} 0.05). The trained model was then applied to the test data
classifying samples as either control or SRN. The VSEARCH Greengenes
method performed slightly worse than the OptiClust \emph{de novo} method
(p-value = 0.030). However the performance on the test data for the
OptiClust \emph{de novo}, OptiFit Greengenes, OptiFit self-reference,
and VSEARCH \emph{de novo} approaches were not significantly different
(p-values \textgreater{} 0.05; Figures 2B and 2C). These results
indicate that new data could be fit to existing OTU clusters using
OptiFit without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification using OTU
relative abundance. A potential problem with using OptiFit is that any
sequences from the new samples that do not map to the existing OTU
clusters will be discarded, resulting in a possible loss of information.
However, we demonstrated that OptiFit can be used to fit new sequence
data into existing OTU clusters and it could perform as well in
predicting SRN compared to \emph{de novo} clustering all the sequence
data together. In this instance, the performance of OptiFit was
equivalent to using a database-reference-based method despite the lower
quality of the OTU clusters in the database-reference-based approach.
This likely indicates that the sequences that are important to the model
are well characterized by the reference database. However, a less well
studied system may not be as well characterized by a reference-database
which would make the ability to utilize one's own data a reference an
exciting possiblility. The ability to integrate data from new samples
into existing OTUs enables the implementation of a single machine
learning model. This is important for model implementation because not
all of the data needs to be available or known at the time of model
generation. A robust machine learning model can be implemented as part
of a non-invasive and low-cost diagnostic for SRN and other diseases.

@@

\textbf{Dataset.} Raw 16S rRNA gene sequence data from the V4 region
were previously generated from human stool samples. Sequences were
downloaded from the NCBI Sequence Read Archive (accession no. SRP062005)
(\protect\hyperlink{ref-baxter2016}{5},
\protect\hyperlink{ref-edgar2011}{9}). This dataset contains stool
samples from 490 subjects. For this analysis, samples from subjects
identified in the metadata as normal, high risk normal, or adenoma were
categorized as ``normal'', while samples from subjects identified as
advanced adenoma or carcinoma were categorized as ``screen relevant
neoplasia'' (SRN). The resulting dataset consisted of 261 normal samples
and 229 SRN samples.

\textbf{Data processing.} The full dataset was preprocessed with mothur
(v1.47) (\protect\hyperlink{ref-schloss2009}{10}) to join forward and
reverse reads, merge duplicate reads, align to the SILVA reference
database (v132) (\protect\hyperlink{ref-quast2013}{11}), precluster,
remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{9}),
assign taxonomy, and remove non-bacterial reads following the Schloss
Lab MiSeq standard operating procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each sample was in the training
set an average of 80 times (standard deviation = 4.1) and the test set
an average of 20 times (standard deviation = 4.1).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  OptiFit Self: The preprocess data was split into the training and
  testing sets. The training set was clustered into OTUs using
  OptiClust, then the test set was fit to the OTUs of the training set
  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}).
  The OptiFit algorithm was run with method open so that any sequences
  that did not map to the existing OTU clusters would form new OTUs. The
  data was then subsampled to 10,000 reads and any novel OTUs from the
  test set were removed. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\item
  OptiFit Greengenes: Reference sequences from the Greengenes database
  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{12}) were downloaded
  and processed with mothur by trimming to the V4 region and clustered
  \emph{de novo} with OptiClust
  (\protect\hyperlink{ref-westcott2017}{6}). The preprocessed data was
  fit to the clustered reference data using OptiFit with the method open
  to allow any sequences that did not map to the existing reference
  clusters would form new OTUs. The data was then subsampled to 10,000
  reads and any novel OTUs from the test set were removed. The dataset
  was then split into two sets where 80\% of the samples were assigned
  to the training set and 20\% to the testing set. This process was
  repeated for each of the 100 splits resulting in 100 training and
  testing datasets.
\item
  VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
  v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) directly to
  unprocessed Greengenes 97\% OTU reference alignment consistent with
  how VSEARCH is typically used by the QIIME2 software for
  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{8}).
  The data was then subsampled to 10,000 reads and any novel OTUs from
  the test set were removed. The dataset was then split into two sets
  where 80\% of the samples were assigned to the training set and 20\%
  to the testing set. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\item
  OptiClust \emph{de novo}: All the preprocessed data was clustered
  together with OptiClust (\protect\hyperlink{ref-westcott2017}{6}) to
  generate OTUs. The data was subsampled to 10,000 reads per sample and
  the resulting abundance tables were split into the training and
  testing sets. The process was repeated for each of the 100 splits
  resulting in 100 training and testing datasets.
\item
  VSEARCH \emph{de novo}: All the preprocessed data was clustered using
  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) with 97\%
  identity and then subsampled to 10,000 reads per sample. The process
  was repeated for each of the 100 splits resulting in 100 training and
  testing datasets for both workflows.
\end{enumerate}

\textbf{\emph{Machine Learning.}} A random forest model was trained with
the R package mikrompl (v 1.2.0)
(\protect\hyperlink{ref-topuxe7uoglu2021}{13}) to predict the diagnosis
(SRN or normal) for the samples in the test set for each data split. The
training set was preprocessed to normalize OTU counts (scale and
center), collapse correlated OTUs, and remove OTUs with zero variance.
The preprocessing from the training set was then applied to the test
set. Any OTUs in the test set that were not in the training set were
removed. P-values comparing model performance were calculated as
previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{14}). The
averaged ROC curves were plotted by taking the average and standard
deviation of the sensitivity at each specificity value.

\textbf{\emph{Code Availability.}}

The analysis workflow was implemented in Snakemake
(\protect\hyperlink{ref-koster2012}{15}). Scripts for analysis were
written in R (\protect\hyperlink{ref-R2020}{16}) and GNU bash
(\protect\hyperlink{ref-GNUbash}{17}). The software used includes mothur
v1.47.0 (\protect\hyperlink{ref-schloss2009}{10}), VSEARCH v2.15.2
(\protect\hyperlink{ref-rognes2016}{7}), RStudio
(\protect\hyperlink{ref-RStudio2019}{18}), the Tidyverse metapackage
(\protect\hyperlink{ref-wickham2019}{19}), R Markdown
(\protect\hyperlink{ref-xie_r_2018}{20}), the SRA toolkit
(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{21}), and conda
(\protect\hyperlink{ref-noauthor_anaconda_2016}{22}). The complete
workflow and supporting files required to reproduce this study are
available at:
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_mBio_2023}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{1}
\leavevmode\vadjust pre{\hypertarget{ref-Morgan2012}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{\textbf{Morgan XC}, \textbf{Tickle TL}, \textbf{Sokol
H}, \textbf{Gevers D}, \textbf{Devaney KL}, \textbf{Ward DV},
\textbf{Reyes JA}, \textbf{Shah SA}, \textbf{LeLeiko N}, \textbf{Snapper
SB}, \textbf{Bousvaros A}, \textbf{Korzenik J}, \textbf{Sands BE},
\textbf{Xavier RJ}, \textbf{Huttenhower C}. 2012. Dysfunction of the
intestinal microbiome in inflammatory bowel disease and treatment.
Genome Biol \textbf{13}:R79.}

\leavevmode\vadjust pre{\hypertarget{ref-Sobhani2011}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{\textbf{Sobhani I}, \textbf{Tap J},
\textbf{Roudot-Thoraval F}, \textbf{Roperch JP}, \textbf{Letulle S},
\textbf{Langella P}, \textbf{Corthier G}, \textbf{Tran Van Nhieu J},
\textbf{Furet JP}. 2011. Microbial dysbiosis in colorectal cancer
({CRC}) patients. PLoS One \textbf{6}:e16393.}

\leavevmode\vadjust pre{\hypertarget{ref-westcott2015}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo
clustering methods outperform reference-based methods for assigning 16S
rRNA gene sequences to operational taxonomic units. PeerJ
\textbf{3}:e1487.
doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.}

>>>>

The ability to use 16S rRNA gene sequence data to train machine learning
classification models offers the opportunity diagnose patients based on
the composition of their microbiome. In some applications the taxonomic
resolution that provides the best models may require the use of \emph{de
novo} OTUs whose composition changes when new data are added. We
previously developed a new reference-based approach, OptiFit, that fits
new sequence data to existing \emph{de novo} OTUs without changing the
composition of the original OTUs. While OptiFit produces OTUs that are
as high quality as \emph{de novo} OTUs, it is unclear whether this
method for fitting new sequence data into existing OTUs will impact the
performance of classification models relative to models trained and
tested only using \emph{de novo} OTUs. We used OptiFit to cluster
sequences into existing OTUs and evaluated model performance in
classifying a dataset containing samples from patients with and without
colonic screen relevant neoplasia (SRN). We compared the performance of
this model to standard methods including \emph{de novo} and
database-reference-based clustering. We found that using OptiFit
performed as well or better in classifying SRNs. OptiFit can streamline
the process of classifying new samples by avoiding the need to retrain
models using reclustered sequences.

@@

There is great potential for using microbiome data to aid in diagnosis.
A challenge with \emph{de novo} OTU-based classification models is that
16S rRNA gene sequences are often assigned to OTUs based on similarity
to other sequences in the dataset. If data are generated from new
patients, the old and new sequences must be reclustered to OTUs and the
classification model retrained. Yet there is a desire to have a single,
validated model that can be widely deployed. To overcome this obstacle,
we applied the OptiFit clustering algorithm to fit new sequence data to
existing OTUs allowing for reuse of the model. A random forest model
implemented using OptiFit performed as well as the traditional reassign
and retrain approach. This result shows that it is possible to train and
apply machine learning models based on OTU relative abundance data that
do not require retraining or the use of a reference database.

@@

There is increasing interest in training machine learning models to
diagnose diseases such as Crohn's disease and colorectal cancer using
the relative abundance of clusters of similar 16S rRNA gene sequences
(\protect\hyperlink{ref-baxter2016}{1},
\protect\hyperlink{ref-gevers2014}{2}). These models have been used to
identify sequence clusters that are important for distinguishing between
individuals from different disease categories
(\protect\hyperlink{ref-duvallet2017}{3}). There is also an opportunity
to train models and apply them to classify samples from new individuals.
For example, a model for colorectal cancer could be trained, ``locked
down'', and applied to samples from new patients.

To apply these models to new samples the composition of the clusters
would need to be independent of the new data. For example, amplicon
sequence variants (ASVs) are defined without consideration of sequences
in other samples, phylotypes are defined by clustering sequences that
have the same taxonomy (e.g., to the same family) when classified using
a taxonomy database, and closed reference operational taxonomic units
(OTUs) are defined by mapping sequences to a collection of reference
OTUs. In contrast, \emph{de novo} approaches cluster sequences based on
their similarity to other sequences in the dataset and can change when
new data are added. Although it would be preferrable to select an
approach that generates stable clusters, there may be cases where OTUs
generated by a \emph{de novo} approach outperform those of the other
taxonomic levels. In fact, we recently trained machine learning models
for classifying patients with and without screen relevant neoplasias
(SRNs) in their colons and found that OTUs generated \emph{de novo}
using the OptiClust algorithm performed better than those generated
using ASVs or at higher taxonomic levels
(\protect\hyperlink{ref-Armour2022}{4}).

It could be possible to construct reference OTUs and map new sequences
to those OTUs to attain similar performance as was seen with the
OptiClust-generated OTUs. The traditional approach to reference-based
clustering of sequences to OTUs has multiple drawbacks and does not
produce clusters as good as those generated using OptiClust
(\protect\hyperlink{ref-sovacool2022}{5}). Sovacool \emph{et al}
recently described OptiFit, a method for fitting new sequence data into
existing OTUs that overcomes the limitations of traditional
reference-based clustering (\protect\hyperlink{ref-sovacool2022}{5}).
OptiFit allows researchers to fit new data into existing OTUs defined
from the same dataset resulting in clusters that are as good as if they
had all been clustered with OptiClust. We tested whether
OptiClust-generated OTUs could be used to train models that were then
used to classify held out samples after clustering their sequences to
the model's OTUs using OptiFit.

To test how the model performance compared between using \emph{de novo}
and reference-based clustering approaches, we used a publicly available
dataset of 16S rRNA gene sequences from stool samples of healthy
subjects (n = 226) as well as subjects with screen-relevant neoplasia
(SRN) consisting of advanced adenoma and carcinoma (n = 229)
(\protect\hyperlink{ref-baxter2016}{1}). For the \emph{de novo}
workflows, the 16S rRNA sequence data from all samples were clustered
into OTUs using the OptiClust algorithm in mothur 1 and the VSEARCH
algorithm used in QIIME2 (\protect\hyperlink{ref-rognes2016}{6},
\protect\hyperlink{ref-bolyen2019}{7}). For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). For traditional reference-based
clustering (database-reference-based), we used OptiFit to fit the
sequence data into OTUs based on the commonly used greengenes reference
database. To compare with another commonly used method, we also used
VSEARCH to map sequences to reference OTUs from the greengenes database
with the parameters used by QIIME2. Again, the data was then split into
training and testing sets, hyperparameters tuned, and performance
evaluated on the testing set (Figure 1B). In the OptiFit self-reference
workflow (self-reference-based), the data was split into a training and
a testing set. The training set was clustered into OTUs and used to
train a classification model. The OptiFit algorithm was used to fit
sequence data of samples not part of the training data into the training
OTUs and classified using the best hyperparameters (Figure 1C). For each
of the workflows the process was repeated for 100 random splits of the
data to account for variation caused by the choice of the random number
generator seed.

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is an
objective metric used to measure OTU cluster quality based on the
similarity of all pairs of sequences and whether they are appropriately
clustered or not (\protect\hyperlink{ref-westcott2015}{8}). We expected
the MCC scores produced by the OptiFit workflow to be similar to that of
\emph{de novo} clustering using the OptiClust algorithm. In the OptiFit
workflow the test data was fit to the clustered training data for each
of the 100 data splits resulting in an MCC score for each split of the
data. In the remaining workflows, the data was only clustered once and
then split into the training and testing sets resulting in a single MCC
score for each method. Indeed, the MCC scores were similar between the
OptiClust \emph{de novo} (MCC = 0.884) and OptiFit self-reference
workflows (average MCC = 0.879, standard deviation = 0.002). Consistent
with prior findings, the reference-based methods produced lower MCC
scores (OptiFit greengenes MCC = 0.786; VSEARCH greengenes MCC = 0.531)
than the \emph{de novo} methods (OptiClust \emph{de novo} MCC = 0.884;
VSEARCH \emph{de novo} MCC = 0.641)
(\protect\hyperlink{ref-sovacool2022}{5}). Another metric we examined
for the OptiFit workflow was the fraction of sequences from the test set
that mapped to the reference OTUs. Since sequences that did not map to
reference OTUs were eliminated, if a high percentage of reads did not
map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average 99.8\% (standard deviation = 0.7\%) of sequences in the
subsampled test set mapped to the reference OTUs. This number is higher
than the average fraction of reads mapped in the OptiFit greengenes
workflow (mean = 96.8\%, standard deviation = 3.5). These results
indicate that the OptiFit self-reference method performed as well as the
OptiClust \emph{de novo} method and better than using an external
database.

We next assessed model performance using OTU relative abundances from
the training data from the workflows to train a model to predict SRNs
and used the model on the held-out data. Using the predicted and actual
diagnosis classification, we calculated the area under the receiver
operating characteristic curve (AUROC) for each data split. During
cross-validation (CV) training, the performance of the OptiFit
self-reference and OptiClust \emph{de novo} models were not
significantly different (p-value = 0.066; Figure 2A), while performance
for both VSEARCH methods was significantly lower than the OptiClust
\emph{de novo}, OptiFit self, and OptiFit greengenes methods (p-values
\textless{} 0.05). The trained model was then applied to the test data
classifying samples as either control or SRN. The VSEARCH greengenes
method performed slightly worse than the OptiClust \emph{de novo} method
(p-value = 0.030). However, the performance on the test data for the
OptiClust \emph{de novo}, OptiFit greengenes, OptiFit self-reference,
and VSEARCH \emph{de novo} approaches were not significantly different
(p-values \textgreater{} 0.05; Figures 2B and 2C). These results
indicate that new data could be fit to existing OTU clusters using
OptiFit without impacting model performance.

Random forest machine learning models trained using OptiClust-generated
OTUs and tested using OptiFit-generated OTUs performed as well as a
model trained using entirely \emph{de novo} OTU assignments. A potential
problem with reference-based clustering methods is that sequences that
do not map to the reference OTUs are discarded, resulting in a possible
loss of information. However, we demonstrated that the training samples
represented the most important OTUs for classifying samples. Missing
important OTUs is more of a risk when using a database-reference-based
method since not all environments are well represented in public
databases. Despite this and the lower quality OTUs, the
database-reference-based approach performed as well as the models
generated using OptiFit. This likely indicates that the sequences that
were important to the model were well characterized by the greengenes
reference OTUs. However, a less well studied system may not be as well
characterized by a reference-database which would make the ability to
utilize one's own data a reference an exciting possibility. Our results
highlight that OptiFit overcomes a significant limitation with machine
learning models trained using \emph{de novo} OTUs. This is an important
result for those applications where models trained using \emph{de novo}
OTUs outperform models generated using methods that produce clusters
that do not depend on which sequences are included in the dataset.

@@

\textbf{Dataset.} Raw 16S rRNA gene sequence data from the V4 region
were previously generated from human stool samples. Sequences were
downloaded from the NCBI Sequence Read Archive (accession no. SRP062005)
(\protect\hyperlink{ref-baxter2016}{1}). This dataset contains stool
samples from 490 subjects. For this analysis, samples from subjects
identified in the metadata as normal, high risk normal, or adenoma were
categorized as ``normal'', while samples from subjects identified as
advanced adenoma or carcinoma were categorized as ``screen relevant
neoplasia'' (SRN). The resulting dataset consisted of 261 normal samples
and 229 SRN samples.

\textbf{Data processing.} The full dataset was preprocessed with mothur
(v1.47) (\protect\hyperlink{ref-schloss2009}{9}) to join forward and
reverse reads, merge duplicate reads, align to the SILVA reference
database (v132) (\protect\hyperlink{ref-quast2013}{10}), precluster,
remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{11}),
assign taxonomy, and remove non-bacterial reads following the Schloss
Lab MiSeq standard operating procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each sample was in the training
set an average of 80 times (standard deviation = 4.1) and the test set
an average of 20 times (standard deviation = 4.1).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  OptiFit Self: The preprocess data was split into the training and
  testing sets. The training set was clustered into OTUs using
  OptiClust, then the test set was fit to the OTUs of the training set
  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{5}).
  The OptiFit algorithm was run with method open so that any sequences
  that did not map to the existing OTU clusters would form new OTUs. The
  data was then subsampled to 10,000 reads and any novel OTUs from the
  test set were removed. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\item
  OptiFit Greengenes: Reference sequences from the Greengenes database
  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{12}) were downloaded
  and processed with mothur by trimming to the V4 region and clustered
  \emph{de novo} with OptiClust

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    The preprocessed data was fit to the clustered reference data using
    OptiFit with the method open to allow any sequences that did not map
    to the existing reference clusters would form new OTUs. The data was
    then subsampled to 10,000 reads and any novel OTUs from the test set
    were removed. The dataset was then split into two sets where 80\% of
    the samples were assigned to the training set and 20\% to the
    testing set. This process was repeated for each of the 100 splits
    resulting in 100 training and testing datasets.
  \end{enumerate}
\item
  VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
  v2.15.2 (\protect\hyperlink{ref-rognes2016}{6}) directly to
  unprocessed Greengenes 97\% OTU reference alignment consistent with
  how VSEARCH is typically used by the QIIME2 software for
  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{7}).
  The data was then subsampled to 10,000 reads and any novel OTUs from
  the test set were removed. The dataset was then split into two sets
  where 80\% of the samples were assigned to the training set and 20\%
  to the testing set. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  OptiClust \emph{de novo}: All the preprocessed data was clustered
  together with OptiClust 1 to generate OTUs. The data was subsampled to
  10,000 reads per sample and the resulting abundance tables were split
  into the training and testing sets. The process was repeated for each
  of the 100 splits resulting in 100 training and testing datasets.
\item
  VSEARCH \emph{de novo}: All the preprocessed data was clustered using
  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{6}) with 97\%
  identity and then subsampled to 10,000 reads per sample. The process
  was repeated for each of the 100 splits resulting in 100 training and
  testing datasets for both workflows.
\end{enumerate}

\textbf{Machine Learning.} A random forest model was trained with the R
package mikrompl (v 1.2.0)
(\protect\hyperlink{ref-topuxe7uoglu2021}{13}) to predict the diagnosis
(SRN or normal) for the samples in the test set for each data split. The
training set was preprocessed to normalize OTU counts (scale and
center), collapse correlated OTUs, and remove OTUs with zero variance.
The preprocessing from the training set was then applied to the test
set. Any OTUs in the test set that were not in the training set were
removed. P-values comparing model performance were calculated as
previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{14}). The
averaged ROC curves were plotted by taking the average and standard
deviation of the sensitivity at each specificity value.

\textbf{Code Availability.} The analysis workflow was implemented in
Snakemake (\protect\hyperlink{ref-koster2012}{15}). Scripts for analysis
were written in R (\protect\hyperlink{ref-R2020}{16}) and GNU bash
(\protect\hyperlink{ref-GNUbash}{17}). The software used includes mothur
v1.47.0 (\protect\hyperlink{ref-schloss2009}{9}), VSEARCH v2.15.2
(\protect\hyperlink{ref-rognes2016}{6}), RStudio
(\protect\hyperlink{ref-RStudio2019}{18}), the Tidyverse metapackage
(\protect\hyperlink{ref-wickham2019}{19}), R Markdown
(\protect\hyperlink{ref-xie_r_2018}{20}), the SRA toolkit
(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{21}), and conda
(\protect\hyperlink{ref-noauthor_anaconda_2016}{22}). The complete
workflow and supporting files required to reproduce this study are
available at:
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2023}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{1}
\leavevmode\vadjust pre{\hypertarget{ref-baxter2016}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
MAM}, \textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.}

\leavevmode\vadjust pre{\hypertarget{ref-gevers2014}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{\textbf{Gevers D}, \textbf{Kugathasan S}, \textbf{Denson
LA}, \textbf{Vázquez-Baeza Y}, \textbf{Van~Treuren W}, \textbf{Ren B},
\textbf{Schwager E}, \textbf{Knights D}, \textbf{Song SJ},
\textbf{Yassour M}, \textbf{Morgan XC}, \textbf{Kostic AD}, \textbf{Luo
C}, \textbf{González A}, \textbf{McDonald D}, \textbf{Haberman Y},
\textbf{Walters T}, \textbf{Baker S}, \textbf{Rosh J}, \textbf{Stephens
M}, \textbf{Heyman M}, \textbf{Markowitz J}, \textbf{Baldassano R},
\textbf{Griffiths A}, \textbf{Sylvester F}, \textbf{Mack D}, \textbf{Kim
S}, \textbf{Crandall W}, \textbf{Hyams J}, \textbf{Huttenhower C},
\textbf{Knight R}, \textbf{Xavier RJ}. 2014. The treatment-naive
microbiome in new-onset crohn{'}s disease. Cell Host \& Microbe
\textbf{15}:382--392.
doi:\href{https://doi.org/10.1016/j.chom.2014.02.005}{10.1016/j.chom.2014.02.005}.}

\leavevmode\vadjust pre{\hypertarget{ref-duvallet2017}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{\textbf{Duvallet C}, \textbf{Gibbons SM}, \textbf{Gurry
T}, \textbf{Irizarry RA}, \textbf{Alm EJ}. 2017. Meta-analysis of gut
microbiome studies identifies disease-specific and shared responses.
Nature Communications \textbf{8}:1784.
doi:\href{https://doi.org/10.1038/s41467-017-01973-8}{10.1038/s41467-017-01973-8}.}
